---
title: "Report: Becoming Father"
subtitle: "Age at first birth among men in Germany based on the SOEP"
author: "Henrik Schubert"
date: "2023-06-15"
output: pdf_document
abstract: "Men's fertility patterns deviate from women's, with a shift towards later ages and a wider age distribution of childbearing. However, limited information exists on the age distribution of first births among men. This study utilizes data from the Socio-ökonomisches Panel (SOEP) to investigate the transition to fatherhood. Non-parametric approaches and survival models are used to explore the impact of age, while considering socio-economic factors. Cohort shifts and East-West disparities are emphasized. This study contributes to the understanding of men's fertility by examining the age distribution of first births. Using SOEP data, insights are gained into the interplay between age, socio-economic factors, and men's fertility. This research aids decision-making on demographic challenges in modern societies."
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")


# Load the functions
source("Functions/packages.R")
source("Functions/functions.R")
source("Functions/graphics.R")


# Load the survival package
library(survival)
library(eha)
library(casebase)
library(GGally)
library(stargazer)
library(survminer)

# Set cohorts
birthyears <- 1945:1995

# Specify the knots for the spline regression
knots <- c(20, 25, 30, 35, 40)

# Switch to TRUE, if you would like to re-estimate the models
estimate <- FALSE
```


# Purpose
Fertility of men deviates from fertility of women. Research points at a wider age-distribution of childbearing and that fertility is more shifted towards the later ages. Despite the increasing evidence on sex differences with respect to age-specific fertility, the information on the age distribution of first births among men remains scarce. For that reason this study utilizes the *Socio-ökonomisches Panel* (SOEP) in order to describe the transition to fatherhood. We use non-parametric approaches as well as survival models to better investigate the effect of age net of other socio-econonmic factors. A focus of this study lies on cohort differences and differences between East and West.


## Data wrangling
For the study we harness the *biobirth* questionnaire from SOEP. The questionnaire contains questions on biological children of the respondent. The Figure \@ref(fig:interview-dates) below illustrates the distribution of interview years for that particular questionnaire. It becomes visible that the interviews were mostly executed after the year 2000 and they were biannually.


```{r cohort-distr, fig.cap = "Distribution of biobirth interviews in the SOEP"}
### Load bio-birth data ---------------------------------------------- 
if(all(isFALSE(estimate) &
       file.exists("Data/spell_data.Rda") &
       file.exists("Data/person_data.Rda"))){
  
  # Load the data
  load("Data/spell_data.Rda")
  load("Data/person_data.Rda")
  
}else{


### Clean the bio-birth data  ----------------------------------------

# Load the birth data
fert <- read_stata("SOEP_V36/Stata/biobirth.dta")

# Remove respondents that where no asked the question
fert <- fert |> filter(bioyear != -1 & gebjahr != -1)

# Filter men
fert <- fert |> filter(sex == 1)

# Filter only relevant cohorts
fert <- fert |> filter(gebjahr %in% birthyears)

# Remove unimportant variables
fert <- fert |> select(!starts_with("kidsex"))

# Make everything as double
fert <- fert |> mutate(across(where(is.factor), as.double))

# Make missing, where values are either -2 or -1
fert <- fert |> replace_with_na_all(condition = ~.x %in% c(-2, -1))

# Clean the names
names(fert) <- sub("(.*)(\\d{2})$", "\\1_\\2", names(fert))

# Make a life-course perspective
fert2 <- fert |> pivot_longer(cols = starts_with("kid"),
                              names_pattern = "([a-z]*)_([0-9]*)",
                              values_to = "Value",
                              names_to = c("Variable", "Number"))

# Filter first births
fert2 <- fert2 |> filter(Number == "01")



# Pivot wider
fert2 <- fert2 |> pivot_wider(names_from = c(Variable, Number),
                              values_from = Value)

# Create cohorts - split by 5 year groups
fert2 <- fert2 |> mutate(cohort = cut(gebjahr, breaks = seq(min(birthyears), max(birthyears), by = 10), include.lowest = TRUE, dig.lab = 4))

# Double check
fert2 <- fert2 |> filter(!is.na(gebjahr) & !is.na(bioyear))

# Create an event and censoring variable
fert2 <- fert2 |> mutate(Event = if_else(is.na(kidgeb_01), 0, 1),
                         Censoring = if_else(Event == 0, bioyear - gebjahr, kidgeb_01 - gebjahr))

# Save the data
save(fert2, file = "Data/person_data.Rda")

}

# Plot the distribution of cohorts
ggplot(fert2, aes(gebjahr, fill = cohort)) +
  geom_histogram(binwidth = 2, colour = "white") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  guides(fill = guide_legend(nrow = 3, byrow = 2))

# Save the cohort distribution
ggsave(last_plot(), filename = "Figures/birthyear_distribution.pdf")
```

As can be derived from the last plot, a 10-year cohort aggregation makes sense. Moreover, we split from 1945 every 10 years.

```{r interview-dates, fig.cap = "Distribution of biobirth interviews in the SOEP"}

### Split the data -----------------------------------------------

# Split the data
spell_data <- survSplit(fert2, cut = 15:55, end = "Censoring", event = "Event", start = "start")

### Save the data
save(spell_data, file = "Data/spell_data.Rda")


### Distribution of questionnaires
ggplot(fert2, aes(bioyear)) +
  geom_histogram() +
  scale_y_continuous(expand = c(0, 0)) +
  ylab("Year of biobirth interview")

```


Table \@ref(table:data-structure1) displays the current shape of the data, when only showing the first 10 cases. Essentially, it is a single spell data set, which includes retrospective information on the fertility history.

```{r data-structure1}
# Make a table of the interview dates
fert2 |>
  arrange(persnr, bioage) |>
  slice_head(n = 10) |>
  select(pid, cohort, bioyear, bioage, Event) |> 
  pander()
```

In \@ref(fig:event-data) illustrates the distribution of censoring or event times across different cohorts. The x-axis of the plot represents the time variable, either the time of the event (first birth) or the time of censoring (such as loss to follow-up or end of the study). The y-axis represents the frequency or proportion of individuals who have experienced the event or remained uncensored at a given time.


This graphical representation provides valuable insights into the survival experience of a population or a specific group, illustrating the probability of experiencing the event at a specific time point. 

```{r event-data}
### Descriptive data ---------------------------------------------

# Plot discriptively
ggplot(subset(fert2, Censoring <= 55), aes(Censoring, fill = as.factor(Event))) +
  geom_density(alpha = 0.3) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  facet_wrap(~ cohort) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(caption = "Data: SOEP Wave36") +
  scale_fill_manual(name = "Birth:", values = c(MPIDRorange, MPIDRblue)) +
  xlab("Age")

# Save
ggsave(last_plot(), filename = "Figures/descriptive_age_firstbirth.pdf")
```


## Survival analysis
As the data exists already in form to proceed with survival analysis, we make some descriptive estimations. First, we estimate kaplan-meier curves using the following estimator: 

$$
\hat{S}(t) = \prod_{t_i \le t} [1  \frac{d_i}{Y_i}] 
$$



### Population
First, we look at the cases which experience the event and which are censored.

```{r}
### Prepare the survival data ------------------------------------

# Look at the survival times
with(fert2, Surv(Censoring, Event))[1:100]

# Make the Kaplan-Meier
km <- survfit(Surv(Censoring, Event) ~ 1, conf.type = "log",
              conf.int = 0.95, type = "kaplan-meier", error = "greenwood",
              data = fert2)

# Plot the kaplan meier
km_result <- with(km, data.frame(time, n.risk, n.event, surv, n.censor, cumhaz, std.chaz, lower, upper)) |> 
  filter(time <= 50) 

# Print the km-table
pander(km_result)

ggplot(km_result, aes(time, y=  surv, ymin = lower, ymax = upper)) +
  geom_line() +
  geom_ribbon(alpha = .3)
```

### Cohort specific
In Figure \@ref(fig:cohort-km), the kaplan-meier curves for specific cohorts are displayed. 

```{r cohort-km}
# Fit by cohort
km_coh <- survfit(Surv(Censoring, Event) ~ cohort, data = fert2,
                  conf.int = 0.95, type = "kaplan-meier", error = "greenwood") 

# Plot
ggsurv(km_coh) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  guides(colour =  guide_legend(nrow = 3, byrow = TRUE))
```

## Smoothed hazard models
Beyond describing the survival process using the Kaplan-Meier estimator, we also estimate smoothed hazard models. The results from the smoothed hazard model are displayed in table. 


```{r}
if(all( isFALSE(estimate) & file.exists("Results/smoothed.Rda"))){
  # Load the model
  load("Results/smoothed.Rda")
  
}else{
  # Run the regression model
mod_cb <- fitSmoothHazard(Event ~ ns(log(Censoring), df = 3) * cohort,
                          data = fert2, 
                          time = "Censoring")

  # Save the model
  save(mod_cb, file = "Results/smoothed.Rda")
}

# Display the results
pander(mod_cb)
```


In order to get a better understanding of the model, I visualized predicted probabilities of first birth by age and cohort in Figure \@ref(fig: pred-smooth).


```{r pred-smooth}
if(all(isFALSE(estimate) & file.exists("Results/predict_smoothed_eha.Rda"))){
  
  # Load the predicted data
  load("Results/predict_smoothed_eha.Rda")
  
}else{
  # Plot the result
plot_results <- plot(mod_cb,
                     hazard.params = list(xvar = "Censoring",
                                          by = "cohort",
                                          alpha = 0.10,
                                          ylab = "Hazard",
                                          plot = FALSE))
  # Save the predicted data
  save(plot_results, file = "Results/predict_smoothed_eha.Rda")
  
}

# Plot the predicted probabilities
plot_results$fit |>
  filter(Censoring >= 15 & Censoring <= 50 ) |> 
  ggplot(aes(Censoring, visregFit, group = cohort, colour = cohort)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 0.2), expand = c(0, 0)) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  ylab("Predicted - age-specific first birth rates") + 
  xlab("Age") +
  ggtitle("Spline logistic regression for first birth among men") +
  labs(caption = "Data: SOEP W36")


```


## Discrete time survival model
While the parametric assumptions allow for more *degrees of freedom*, misspecification of the process may occur. In order to circumvent this issue, we have also estimated discrete time hazard models with splines for the age variables. We set the knots at 5-year age intervals. For this estimation, we use data in long-format, as is illustrated below.


```{r spell-data}
if(all(isFALSE(estimate) & file.exists("Results/discrete_eha_splines.Rda"))){
  
  # Load the data
  load("Results/discrete_eha_splines.Rda")
  
}else{

### Discrete time model ------------------------------------------

# Estimate a logistic regression
logist <- glm(Event ~ ns(Censoring, knots = knots) * cohort, data = spell_data)

# Save the results
save(logist, file = "Results/discrete_eha_splines.Rda")

}

# Create the prediction data
pred_data <- expand.grid(Censoring = 15:55, cohort = unique(fert2$cohort))

# Predict the results
pred_data$prediction <- predict(logist, pred_data)

# Select the data
pred_data <- subset(pred_data, Censoring >= 18 )

# De-select data
pred_data <- pred_data |> filter((cohort == "(1975,1985]" & Censoring <= 35) |
                                   (cohort == "(1985,1995]" & Censoring <= 25 ) |
                                   cohort %in% c("(1945,1955]", "(1955,1965]", "(1965,1975]"))


# Print the spell data
spell_data |>
  arrange(pid, start) |> 
  select(persnr, cohort, start, sumkids, Censoring, Event, kidgeb_01) |> 
  slice_head(n = 15) |>
  pander()
```


The results for the discrete-time logistic regression in form of predicted probabilities are displayed below.


```{r discrete-eha}
# Plot the result
ggplot(pred_data, aes(Censoring, prediction, colour = cohort, group = cohort)) +
  geom_line(size = 1.3)  +
  scale_y_continuous(limits = c(0, 0.15), expand = c(0, 0)) +
  ylab("Predicted - age-specific first birth rates") + 
  xlab("Age") +
  ggtitle("Spline logistic regression for first birth among men") +
  labs(caption = "Data: SOEP W36") +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) 
  

# Save the file
ggsave(last_plot(), filename = "Figures/logistic_splines_soep.pdf")



```


### A non-parametric approach
While the models are useful for incorporating covariates, they may rely on too restrictive assumptions. Therefore, we also used a non-parametric approach to estimate age-specific first birth rates.


We used the spell data and aggregated the exposures as well as the births by age. Than, we simply estimated the rates in the following way:


$$
rate(x) = \frac{B_{firstbirth}(x)}{P_{childless}}
$$

Figure \@ref{fig:plot_raw} illustrates the raw age-specific first birth rates for different cohorts. Because the data is from a survey, the rates show an erratic pattern. Nonetheless, the expected bell-shape becomes apparent.


```{r plot_raw, fig.width = 5, fig.height = 10}

# Estimate the exposures
exposures <- spell_data |> group_by(start, cohort) |> count()

# Count the events
births <- spell_data |> group_by(start, cohort) |> summarise(birth = sum(Event))

# Combine
unparametric <- inner_join(exposures, births) |> mutate(rate = birth / n)


# De-select data
pred_data <- unparametric |> filter((cohort == "(1975,1985]" & start <= 35) |
                                   (cohort == "(1985,1995]" & start <= 25 ) |
                                   cohort %in% c("(1945,1955]", "(1955,1965]", "(1965,1975]"))

# Plot the result
plot_raw <- unparametric |> 
  filter( start >= 18) |> 
  ggplot(aes(start, rate, colour = cohort, group = cohort, shape = cohort)) +
    geom_line() +
    geom_point() +
    facet_wrap( ~ cohort) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 0.15)) +
    ggtitle("Age-specific first-birth rates for men") +
    labs(caption = "Data: SOEP Wave 36") +
    ylab("Age-specific fertility rate (Parity 1)") +
    xlab("Age") +
    guides(colour = guide_legend(nrow = 2, byrow = TRUE))


# Plot the result
plot_raw

```


In order to reduce the noise and random fluctuations, which result from limited case numbers and the spread of the interview dates, we have smoothed the age-specific first birth rates using a *locally estimated scatterplot smoothing* (loess). The results are presented in Figure \@ref{fig:smoothed-rates}


```{r smoothed-rates}

# Plot interpolated
plot_interpol <- unparametric |>
  filter(start >= 18) |> 
  ggplot(aes(start, rate, colour = cohort, group = cohort, linetype = cohort, fill = cohort)) +
    geom_smooth(se = FALSE) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 0.17)) +
    facet_wrap( ~ cohort) +
    ggtitle("Age-specific first-birth rates for men (smoothed)") +
    labs(caption = "Data: SOEP Wave 36") +
    ylab("Age-specific fertility rate (Parity 1)") +
    xlab("Age") +
    guides(colour = guide_legend(nrow = 2, byrow = TRUE),
           linetype = guide_legend(nrow = 2, byrow = TRUE)) +
    scale_linetype_manual(values = c("dashed", "dotted", "longdash", "twodash", "solid"))

# Plot the result
plot_interpol
```


# Comparison by birth region
It is very likely that some of the change in the age distribution is driven by the impact reunification, which caused migration as well as fertility postponement. Thus, we estimated non-parametric age-specific first birth rates separately by birth region. The sample was split into persons who were born in East-Germany and respondents who were born in West Germany. Following common practice, respondents from Berlin were classified as East-German.


```{r}
if(all(isFALSE(estimate) & file.exists("Data/region_spell_data.Rda"))){
  
  # Load the data
  load("Data/region_spell_data.Rda")
  
  
}else{
  
### Prepare the background data ---------------------------

# Load the data
id <- read_dta(file = "SOEP_V36/Stata/ppfad.dta")

# Select variables
id <- subset(id, select = c(persnr, pid, birthregion, loc1989))

#

# Clean the birthregion
id$birthregion <- ifelse(id$birthregion %in% 11:16, "East",
       ifelse(id$birthregion %in% 1:10, "West", NA_character_))


### Combine with background variables ----------------------------

# Join with birth region
fert2 <- left_join(fert2, id)

# Plot the share of missing values in birth region
ggplot(fert2, aes(x = birthregion, fill = birthregion)) +
  stat_count() +
  scale_y_continuous(expand = c(0, 0))

# Save the plot
ggsave(last_plot(), filename = "Figures/missing_birthregion.pdf")

# Filter respondents where the birth information are existent
fert2 <- fert2 |> filter(!is.na(birthregion))

# Create spell data
spell_data_reg <- survSplit(fert2, cut = 15:55, end = "Censoring", event = "Event", start = "start")

# Save the data
save(spell_data_reg, file = "Data/spell_data_reg.Rda")

}

# Create the prediction data
pred_data <- expand.grid(Censoring = 15:55, cohort = unique(fert2$cohort), birthregion = c("East", "West"))
```


### Kaplan-Meier by birthregion
Once we have prepared the data, we estimate Kaplan-Meier curves by region.


```{r}
# Fit by cohort
km_coh_reg <- survfit(Surv(Censoring, Event) ~ cohort + birthregion,
                      data = fert2, conf.int = 0.95,
                      type = "kaplan-meier", error = "greenwood") 

# Transform into a data frame
km_coh_reg_data <- surv_summary(km_coh_reg, data = fert2) |> 
  filter(time <= 50)
    
# Plot
ggplot(km_coh_reg_data, aes(x = time, y = surv, colour = birthregion, group = birthregion)) +
  geom_step() +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  guides(colour =  guide_legend(nrow = 3, byrow = TRUE)) +
  facet_wrap( ~ cohort)

# Save the plot
ggsave(last_plot(), filename = "Figures/km_reg-coh.pdf")
  
```


In the next step, we estimate a discrete time survival regression with knots in 5-year intervals, with interactions between cohort and birth region. We than plot the predicted probabilities from the model in \@ref(fig:pred-reg)


```{r pred-reg,  fig.height = 10, fig.width = 5}
# Estimate a logistic regression
logist <- glm(Event ~ ns(Censoring, knots = knots) * cohort * birthregion, 
              data = spell_data_reg)

# Predict the results
pred_data$prediction <- predict(logist, pred_data)

# Select the data
pred_data <- subset(pred_data, Censoring >= 18 )

# De-select data
pred_data <- pred_data |> 
 filter((cohort == "(1975,1985]" & Censoring <= 35) |
       (cohort == "(1985,1995]" & Censoring <= 25 ) |
       cohort %in% c("(1945,1955]", "(1955,1965]", "(1965,1975]"))

# Plot the result
ggplot(pred_data, aes(Censoring, prediction, colour = birthregion, group = birthregion)) +
  geom_line(size = 1.3)  +
  scale_y_continuous(limits = c(0, 0.2), expand = c(0, 0)) +
  ylab("Predicted - age-specific first birth rates") + 
  xlab("Age") +
  facet_wrap( ~ cohort) +
  ggtitle("Spline logistic regression for first birth among men") +
  labs(caption = "Data: SOEP W36") +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE))  


# Save the file
ggsave(last_plot(), filename = "Figures/logistic_reg_soep.pdf")
```


As outlined earlier, the models may suffer from subjectivity and parametric assumptions, while they increase the degrees of freedom. We estimate the age-specific first birth rates using the non-parametric approach as well. The results with the raw birth rates is displayed in Figures \@ref(fig:nonpara-reg).


```{r nonpara-reg,  fig.height = 8, fig.width = 8}
### Unparametric by birthregion ------------------------------------

# Estimate the exposures
exposures <- spell_data_reg |> group_by(start, cohort, birthregion) |> count()

# Count the events
births <- spell_data_reg |> group_by(start, cohort, birthregion) |> summarise(birth = sum(Event))

# Combine
unparametric_reg <- inner_join(exposures, births) |> mutate(rate = birth / n)

# De-select data
unparametric_reg <- unparametric_reg |> 
  filter((cohort == "(1975,1985]" & start <= 35) |
         (cohort == "(1985,1995]" & start <= 25 ) |
         cohort %in% c("(1945,1955]", "(1955,1965]", "(1965,1975]"))

# Plot the result
plot_raw_reg <- unparametric_reg |>
  filter(start >= 18) |> 
  ggplot(aes(start, rate, colour = birthregion, group = birthregion)) +
    geom_line() +
    geom_point() +
    facet_wrap( ~ cohort) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 0.2)) +
    ggtitle("Age-specific first-birth rates for men") +
    labs(caption = "Data: SOEP Wave 36") +
    ylab("Age-specific fertility rate (Parity 1)") +
    xlab("Age") 


# Print the result
plot_raw_reg
```


Again, we used *loess* to smooth the rates and to yield a more schematic result. The result is displayed in Figure \@ref(fig:nonpara-smooth-reg).


```{r nonpara-smooth-reg, fig.height = 8, fig.width = 8}
# Plot interpolated
plot_interpol_reg <- unparametric_reg |>
  filter(start >= 18) |> 
  ggplot(aes(start, rate, colour = birthregion, group = birthregion)) +
    geom_smooth(se = FALSE) +
    facet_wrap( ~ cohort) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 0.2)) +
    ggtitle("Age-specific first-birth rates for men (smoothed)") +
    labs(caption = "Data: SOEP Wave 36") +
    ylab("Age-specific fertility rate (Parity 1)") +
    xlab("Age") 

# Plot the interpolated result
plot_interpol_reg
```



## Parametric regression models


To allow for the inclusion of covariates, we used parametric event-history models.
In order to abstain from too restrictive assumptions regarding the parametric shape, we have estimated models with several parametric specifications and compared the results using log-rank tests.


### Exponential model

```{r parametric-models1, results='asis'}
### Make parametric hazard models --------------------------------

# Exponential
exp <- par_surv(distribution = "exponential")
stargazer(exp, header = FALSE, type = 'latex')
```

### Weibull model
```{r parametric-models2, results='asis'}
# Weibull
weib <- par_surv(distribution = "weibull")
stargazer(weib, header = FALSE, type = 'latex')
```

### Gaussian model
```{r parametric-models3, results='asis'}
# Gompertz
#gomp <- par_surv(distribution = "gompertz")

# Gaussian
gauss <- par_surv(distribution = "gaussian")
stargazer(gauss, header = FALSE, type = 'latex')
```

### Log-normal model
```{r parametric-models4, results='asis'}
# Lognormal
lognor <- par_surv(distribution = "lognormal")
stargazer(gauss, header = FALSE, type = 'latex')
```

### Log-logistic model
```{r parametric-models5, results='asis'}
# Log-logistic
loglog <- par_surv(distribution = "loglogistic")
stargazer(loglog, header = FALSE, type = 'latex')

```

